DOpElib:\\ Differential Equations and Optimization Environment

### 2017-08-07 Examples/Documentation/introduction.tex
Updated Documentation

\section{License information}
Copyright (C) 2012--2016 by the DOpElib authors\\[2mm]
%
This file is part of DOpElib\\[2mm]
%
DOpElib is free software: you can redistribute it
and/or modify it under the terms of the GNU General Public
License as published by the Free Software Foundation, either
version 3 of the License, or (at your option) any later
version.\\[2mm]
%
DOpElib is distributed in the hope that it will be
useful, but WITHOUT ANY WARRANTY; without even the implied
warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
PURPOSE.  See the GNU General Public License for more
details.\\[2mm]
%
Please refer to the file LICENSE.TXT included in this distribution
for further information on this license.

\newpage
\section{References}
If you like the library and use it for your own projects, please give credits by 
referencing the project \dope{}~\cite{dope} using the following bibtex entry:

\begin{lstlisting}
@MISC{dope,
  key   = {DOpElib},
  title = {The {D}ifferential {E}quation and 
               {O}ptimization {E}nvironment: \textsc{DOpElib}},
  url   = {http://www.dopelib.net},
  note  = {\texttt{http://www.dopelib.net}}
}
\end{lstlisting}

@@

Under MAC OSX, you need to install the \texttt{XCode} tools delivered with the operating system 
or available for free \texttt{developer.apple.com/xcode}. Unfortunately, the delivered 
version of \texttt{g++} is too old, so you need to install the real thing. To do so, 
download \texttt{MacPorts} from \texttt{macports.org}. Once you have installed MacPorts 
you can use it to install additional Linux software. 
First, update the MacPorts installation \texttt{sudo port selfupdate}
after that you can install a new version of \texttt{g++} using for instance 
\texttt{sudo port install gcc48} to install version $4.8$ of the compiler.
Afterwards, you need to set the search path appropriate to find the macports version
of \texttt{g++}, to check if this has been done use \texttt{g++ -v}.

@@

\subsubsection{SNOPT}
If you would like to use the features offered in our SNOPT wrapper. You will 
need to obtain a license for \texttt{SNOPT} 
\url{http://www.sbsi-sol-optimize.com/asp/sol_product_snopt.htm}.
Unfortunately this is at present not available for free, but you should 
check if 
there is a department license already available.
For further information you should consult the file 
\texttt{ThirdPartyLibs/SNOPT.INSTALLNOTES}. In particular you need to configure 
\texttt{deal.II} with at least with enabled UMFPACK.

@@

The installation is straighforward and has been tested on openSUSE 12.1
machines as well as on MAC. At the end of the installtion do not forget to
add ipopt to your \texttt{LD\underline{ }LIBRARY\underline{ }PATH}:
\begin{lstlisting}
**************************************************************
                 Installation complete!
Add /home/..../dopelib-4.0/ThirdPartyLibs/ipopt/lib64
    to your LD\_LIBRARY\_PATH variable
**************************************************************
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\section{Installation}\label{sec_installation}
To work with DOpElib, your need necessarily deal.II, which installation 
we describe first. Afterwards, the DOpElib installation is described and finally
other optional packages might be installed.
\begin{itemize}
\item Install \texttt{deal.II} to your home directory, i.e., it should be located in \\
\texttt{\textasciitilde/deal.II}. \\[1mm]
If you prefer another position you can install it anywhere but need 
to set the \texttt{DEAL\_II\_DIR} environment variable, so that it 
will be found by cmake, i.e.,
\begin{lstlisting}
export DEAL_II_DIR=[path to deal.ii]
\end{lstlisting}
\begin{remark}
\textbf{Be careful} if you work in several terminal windows and have several deal.II versions. In this case, the
export-command might have to be done again in that new window. Otherwise
another deal.II version might be linked resulting in linker errors or
segmentation faults without useful error message. In order to check to which
deal.II version DOpE is linked, type 
\begin{lstlisting}
echo $DEAL_II_DIR
\end{lstlisting}
in the current terminal window.
\end{remark}
Detailed installing instructions for deal.II (here the last version
8.4.0) can be found on 
\texttt{http://www.dealii.org/8.4.0/readme.html}. The deal.II 
installation instructions are descriptive enough and we omit 
any further comments and refer to their webpage.
%

%
\item Get a copy of \texttt{DOpElib}, see Section~\ref{sec:obtain} for details.
%
\item Unpack in your preferred directory:
\begin{lstlisting}
~/Software/> tar -xzf dopelib-4.0.tar.gz
\end{lstlisting}
to get
\begin{lstlisting}
~/Software/dopelib-4.0
\end{lstlisting}
To build the \texttt{DOpElib} you have to change to the directory 
\texttt{DOpEsrc} where you can call 
\begin{lstlisting}
~/Software/dopelib-4.0/DOpEsrc> make c-all
\end{lstlisting}
to build and configure the library using cmake. 
You also get various installation options by just typing 
\texttt{make}.
\item To generate documentation, please go into:
\begin{lstlisting}
~/Software/dopelib-4.0/Examples
\end{lstlisting}
Here, you get all options by typing:
\begin{lstlisting}
make
\end{lstlisting}
Specifically, you can create and run all examples, running a test suite,
and generating documentation:
{\footnotesize
\begin{lstlisting}
===========================================================================
=              Makefile for the DOpE documentation                        =
===========================================================================
=                                                                         =
= The following targets exist:                                            =
=    c-all     :  Make all examples using cmake                           =
=    clean     :  Cleaning up all examples                                =
=    tests     :  Run all test param data.                                =
=    c-cat     :  Run clean, run c-all, run tests (combine these commands)=
=    doc       :  Create documentation in pdf file format via latexmk     =
=    distclean :  Cleaning up, including documentation                    =
=    warncheck :  Checks whether all Examples compile without warnings    =
===========================================================================
\end{lstlisting}
}
The options are $c-all$ and $c-cat$ utilizing the cmake build system.

It is \textbf{important} to generate the examples via make c-all 
in the home folder of the examples and \textbf{not} as in deal.II by 
using cmake . in the specific example itself. With this you
would currently destroy the delivered Makefile. See also the FAQ below.

For html documentation, change into:
\begin{lstlisting}
~/Software/dopelib-4.0/doxygen
\end{lstlisting}
and then typing \texttt{make}. Doxygen documentation
will require either 
latex or doxygen to be installed on your computer.
\item If you wish to test if everything worked. To do so you can 
change to the \texttt{Examples} directory and \texttt{make tests} which will give you a 
list of all the examples and whether they behave as expected by the library, see also 
Chapter~\ref{chap:test}.
\item  If you want to use some of the supported third party libraries install them and follow 
 the instructions in \texttt{ThirdPartyLibs/README}. There may be further information 
 in some \texttt{ThirdPartyLibs/*.INSTALLNOTES} that you may want to
 consider.\\[3mm]
 As example, the installation of ipopt works as follows:\\
In the path \texttt{~/dope/ThirdPartyLibs>} type in the terminal\\
\texttt{./install-free-libs.sh}

@@

\subsection*{3.) Problem using make c-all in the Examples folder :}

\begin{lstlisting}
wick@linux:~/Software/dopelib/Examples> make c-all
cd OPT/StatPDE/Example1; make -s c-all
make[2]: *** No targets specified and no makefile found.  Stop.
../../../..//Examples/Make.global_options:2: recipe for target 'c-all' failed
make[1]: *** [c-all] Error 2
Makefile:24: recipe for target 'OPT/StatPDE/Example1' failed
make: *** [OPT/StatPDE/Example1] Error 2
\end{lstlisting}

This means that the Makefile in the specific examples folder has been 
modified in some illegal way. Remove this Makefile 
\begin{lstlisting}
rm Makefile
rm -r CMakeFiles
rm CMakeCache.txt
rm -r autobuild
\end{lstlisting}
and get the original 
one from DOpElib:
\begin{lstlisting}
git checkout Makefile
\end{lstlisting}
Then try again make c-all (make always sure that before 
typing make c-all that the deal.II path is exported - see above).

\subsection*{4.) Different g++ compilers are used for compiling DOpElib and the Examples folder}
\begin{lstlisting}
bernhard@ebook:~/Schreibtisch/Prog/DOpe/Dope/dopelib/Examples> make c-all
cd OPT/StatPDE/Example1; make -s c-all
[ 50%] Building CXX object CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o
c++: error: unrecognized command line option '-fopenmp-simd'
c++: error: unrecognized command line option '-std=c++14'
CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/build.make:62: recipe for target 'CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o' failed
make[4]: *** [CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o] Error 1
CMakeFiles/Makefile2:131: recipe for target 'CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/all' failed
make[3]: *** [CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/all] Error 2
Makefile:83: recipe for target 'all' failed
make[2]: *** [all] Error 2
../../../..//Examples/Make.global_options:2: recipe for target 'c-all' failed
make[1]: *** [c-all] Error 2
Makefile:24: recipe for target 'OPT/StatPDE/Example1' failed
make: *** [OPT/StatPDE/Example1] Error 2
\end{lstlisting}

The reason for this problem might be that the variables c++ and g++ as well as cc and gcc do not match.
\newline
You can remove the old Examples folder and replace it by a new one and set the compilers manually by
\begin{lstlisting}
export CXX=path/to/g++-compiler
export CC=path/to/gcc-compiler
\end{lstlisting}
where path/to/compiler should be the path to the compiler, DOpElib was compiled with.
You can check the current version of your gcc- and g++-compiler with 
\begin{lstlisting}
gcc --version
g++ --version
\end{lstlisting}
Alternatively you can set the variable c++ to your g++ and the variable cc to gcc using update-alternatives
as for example
\begin{lstlisting}
sudo update-alternatives --install /usr/bin/c++ c++ /usr/bin/g++-6 60
sudo update-alternatives --install /usr/bin/cc cc /usr/bin/gcc-6 60
\end{lstlisting}
Thanks to Florian Tischler (Admin at RICAM Linz) for the help.

>>>>

\section{License information}
Copyright (C) 2012--2017 by the DOpElib authors\\[2mm]
%
This file is part of DOpElib\\[2mm]
%
DOpElib is free software: you can redistribute it
and/or modify it under the terms of the GNU General Public
License as published by the Free Software Foundation, either
version 3 of the License, or (at your option) any later
version.\\[2mm]
%
DOpElib is distributed in the hope that it will be
useful, but WITHOUT ANY WARRANTY; without even the implied
warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
PURPOSE.  See the GNU General Public License for more
details.\\[2mm]
%
Please refer to the file LICENSE.TXT included in this distribution
for further information on this license.

\newpage
\section{References}
If you like the library and use it for your own projects, please give credits by 
referencing the project \dope{}~\cite{dope} using the following bibtex entries:

\begin{lstlisting}
@Article{dopelib,
  Title   = {{DOpElib}: {D}ifferential Equations and 
    {O}ptimization {E}nvironment; {A} Goal Oriented 
    Software Library for Solving {PDE}s and Optimization 
    Problems with {PDE}s},
  Author  = {C. Goll and T. Wick and W. Wollner},
  Journal = {Archive of Numerical Software},
  Year    = {2017},
  Number  = {2},
  Pages   = {1--14},
  Volume  = {5},
  Doi     = {10.11588/ans.2017.2.11815}
}

@MISC{dope,
  key   = {DOpElib},
  title = {The {D}ifferential {E}quation and 
               {O}ptimization {E}nvironment: \textsc{DOpElib}},
  url   = {http://www.dopelib.net},
  note  = {\texttt{http://www.dopelib.net}}
}
\end{lstlisting}

@@

Under MAC OSX, you need to install the \texttt{XCode} tools delivered with the operating system 
or available for free \texttt{developer.apple.com/xcode}. Unfortunately, the delivered 
version of \texttt{g++} is too old, so you need to install the real thing. To do so, 
download, e.g., \texttt{MacPorts} from \texttt{macports.org}. Once you have installed MacPorts 
you can use it to install additional Linux software. 
First, update the MacPorts installation \texttt{sudo port selfupdate}
after that you can install a new version of \texttt{g++} using for instance 
\texttt{sudo port install gcc48} to install version $4.8$ of the compiler.
Afterwards, you need to set the search path appropriate to find the macports version
of \texttt{g++}, to check if this has been done use \texttt{g++ -v}.

@@

\subsubsection{SNOPT}
If you would like to use the features offered in our SNOPT wrapper. You will 
need to obtain a license for \texttt{SNOPT} 
\url{http://www.sbsi-sol-optimize.com/asp/sol_product_snopt.htm}.
Unfortunately this is at present not available for free, but you should 
check if 
there is a department license already available.
For further information you should consult the file 
\texttt{ThirdPartyLibs/SNOPT.INSTALLNOTES}. In particular you need to configure 
\texttt{deal.II} at least with enabled UMFPACK.

@@

The installation is straightforward and has been tested on OpenSUSE
12.1, Ubuntu 14.04 LTS and 16.04 LTS
machines as well as on MAC. At the end of the installation do not forget to
add \texttt{IPOPT} to your \texttt{LD\underline{ }LIBRARY\underline{ }PATH}:
\begin{lstlisting}
**************************************************************
                 Installation complete!
Add /home/..../dopelib-4.0/ThirdPartyLibs/ipopt/lib64
    to your LD\_LIBRARY\_PATH variable
**************************************************************
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\section{Installation}\label{sec_installation}
To work with DOpElib, your need necessarily deal.II, which installation 
we describe first. Afterwards, the DOpElib installation is described and finally
other optional packages might be installed.
\begin{itemize}
\item We automatically search for \texttt{deal.II} to your home
  directory, i.e., if \texttt{deal.II} is located in \\
\texttt{\textasciitilde/deal.II} you don't have to do anything. 
\item Alternatively, if you prefer another position you can install it anywhere but need 
to set the \texttt{DEAL\_II\_DIR} environment variable, so that it 
will be found by cmake, i.e.,
\begin{lstlisting}
export DEAL_II_DIR=[path to deal.ii]
\end{lstlisting}
\begin{remark}
\textbf{Be careful} if you work in several terminal windows and have several deal.II versions. In this case, the
export-command might have to be done again in that new window. Otherwise
another deal.II version might be linked resulting in linker errors or
segmentation faults without useful error message. In order to check to which
deal.II version DOpE is linked, type 
\begin{lstlisting}
echo ${DEAL_II_DIR}
\end{lstlisting}
in the current terminal window.
\end{remark}
Detailed installing instructions for deal.II (here the last version
8.4.0) can be found on 
\texttt{http://www.dealii.org/8.4.0/readme.html}. The deal.II 
installation instructions are descriptive enough and we omit 
any further comments and refer to their webpage.
%

%
\item Get a copy of \texttt{DOpElib}, see Section~\ref{sec:obtain} for details.
%
\item If you obtained a tar-ball, unpack in your preferred directory:
\begin{lstlisting}
 tar -xzf dopelib-4.0.tar.gz
\end{lstlisting}
to get a directory 
\begin{lstlisting}
dopelib-4.0/
\end{lstlisting}
containing the source files. If you cloned the git repository you
already have a directory; \texttt{dopelib} by default, containing the sources.
\item Change to the directory with the sources, and then to the
  subdirectory \texttt{DOpEsrc}.
Once there, to build \texttt{DOpElib} you can call 
\begin{lstlisting}
 make c-all
\end{lstlisting}
to build and configure the library using cmake. 
You also get various alternatives by just typing 
\texttt{make}, including the generation of the documentation.
Building the documentation will require  
latex or doxygen to be installed on your computer.
\item If your would like to check the examples, change to the
  subfolder \texttt{Examples} from the main source directory.
Typing 
\begin{lstlisting}
make
\end{lstlisting}
will again provide you with the available options:
{\footnotesize
\begin{lstlisting}
===========================================================================
=              Makefile for the DOpE documentation                        =
===========================================================================
=                                                                         =
= The following targets exist:                                            =
=    c-all     :  Make all examples using cmake                           =
=    clean     :  Cleaning up all examples                                =
=    tests     :  Run all test param data.                                =
=    c-cat     :  Run clean, run c-all, run tests (combine these commands)=
=    doc       :  Create documentation in pdf file format via latexmk     =
=    distclean :  Cleaning up, including documentation                    =
=    warncheck :  Checks whether all Examples compile without warnings    =
===========================================================================
\end{lstlisting}
}
The options are $c-all$ and $c-cat$ utilizing the cmake build system
to create the executables (and optionally run the tests if all went well).

It is \textbf{important} to generate the examples via make c-all 
in the home folder of the examples and \textbf{not} as in deal.II by 
using cmake in the specific example itself. With this you
would currently destroy the delivered Makefile; which is fine if you
don't intend to use it but some automated tests will fail. See also the FAQ below.

\item If you wish to test if everything worked. To do so you can 
change to the \texttt{Examples} directory and \texttt{make tests} which will give you a 
list of all the examples and whether they behave as expected by the library, see also 
Chapter~\ref{chap:test}.
\item  If you want to use some of the supported third party libraries install them and follow 
 the instructions in \texttt{ThirdPartyLibs/README}. There may be further information 
 in some \texttt{ThirdPartyLibs/*.INSTALLNOTES} that you may want to
 consider.\\[3mm]
 As example, the installation of \texttt{IPOPT} works as follows:\\
In the path \texttt{ThirdPartyLibs>} type in the terminal\\
\texttt{./install-free-libs.sh}

@@

\subsection*{3.) Problem using make c-all in the Examples folder :}
Sometimes people encounter problems when calling \texttt{make c-all}. 
This can be due to several reasons:
\begin{itemize}
\item If you abort the script for the c-make auto-configuration prematurely,
you might encounter the following error:
{\small
\begin{lstlisting}
[...]> make c-all
cd OPT/StatPDE/Example1; make -s c-all
make[2]: *** No targets specified and no makefile found.  Stop.
make[1]: *** [c-all] Error 2
make: *** [OPT/StatPDE/Example1] Error 2
\end{lstlisting}
}
This is due to the fact, that some unfinished files from the previous
run are in the way of the script. Go to the corresponding examples
folder, and delete the folder \texttt{autobuild}.

\item Some times the \texttt{Makefiles} in the example folders are
  damaged, e.g., if an in-source cmake build was attempted in the
  example folders. This gives an error message 
{\small
\begin{lstlisting}
[...]> make c-all
cd OPT/StatPDE/Example1; make -s c-all
make[1]: *** No rule to make target `c-all'.  Stop.
make: *** [OPT/StatPDE/Example1] Error 2   
\end{lstlisting}
}
If this happens, go to the corresponding example directory, and replace
the corresponding Makefile from a clean source, e.g., by a new
checkout of the Makefile if you are using the git version.
\end{itemize}

Then try again make c-all (make always sure that before 
typing make c-all that the deal.II path is exported - see above).

\subsection*{4.) Different g++ compilers are used for compiling
  DOpElib and the Examples folder}
On OpenSUSE and Fedora systems we have observed the following error 
when calling \texttt{make c-all} in the examples folder:
{\footnotesize
\begin{lstlisting}
[...]> make c-all
cd OPT/StatPDE/Example1; make -s c-all
[ 50%] Building CXX object CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o
c++: error: unrecognized command line option '-fopenmp-simd'
c++: error: unrecognized command line option '-std=c++14'
CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/build.make:62: 
    recipe for target 'CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o' failed
make[4]: *** [CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/main.cc.o] Error 1
CMakeFiles/Makefile2:131: recipe for target 
    'CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/all' failed
make[3]: *** [CMakeFiles/DOpE-OPT-StatPDE-Example1.dir/all] Error 2
Makefile:83: recipe for target 'all' failed
make[2]: *** [all] Error 2
../../../..//Examples/Make.global_options:2: recipe for target 'c-all' failed
make[1]: *** [c-all] Error 2
Makefile:24: recipe for target 'OPT/StatPDE/Example1' failed
make: *** [OPT/StatPDE/Example1] Error 2
\end{lstlisting}
}

The reason for this problem might be that the variables c++ and g++ as
well as cc and gcc used for compiling \texttt{deal.II} and
\texttt{DOpElib} do not match.
\newline
You can remove the old Examples/autobuild folder and set the compilers manually by
\begin{lstlisting}
export CXX=path/to/g++-compiler
export CC=path/to/gcc-compiler
\end{lstlisting}
where path/to/compiler should be the path to the compiler, DOpElib
(and deal.II) was compiled with.
You can check the current version of your gcc- and g++-compiler with 
\begin{lstlisting}
gcc --version
g++ --version
\end{lstlisting}
It is even better if you automatically set these variables; e.g.,
using your \texttt{\${HOME}.profile} file.

Alternatively you can set the variable c++ to your g++ and the variable cc to gcc using update-alternatives
as for example
\footnote{Thanks to Florian Tischler (Admin at RICAM Linz) for the
  help.}
{\small
\begin{lstlisting}
sudo update-alternatives --install /usr/bin/c++ c++ /usr/bin/g++-6 60
sudo update-alternatives --install /usr/bin/cc cc /usr/bin/gcc-6 60
\end{lstlisting}}

\subsection*{5.) Different g++, cc, cxx compilers are used for compiling
  DOpElib and the Examples folder}

Similar to the previous error, using mpi-compilers bears a similar
problem if the needed libraries are not in your \texttt{\${PATH}}.
Again this problem seems to appear on OpenSUSE and Fedora systems
where this happens by default.
When running \texttt{make c-all} this gives a message like 
{\small
\begin{lstlisting}
Linking CXX executable ../DOpE-PDE-StatPDE-Example1
/usr/[...]/x86_64-suse-linux/bin/ld: cannot find -lmpi_usempi
/usr/[...]/x86_64-suse-linux/bin/ld: cannot find -lmpi_mpifh
/usr/[...]/x86_64-suse-linux/bin/ld: cannot find -lmpi
collect2: error: ld returned 1 exit status
\end{lstlisting}
}

To fix this issue, you need to set the correct environment variables.
We assume you have a standard installation of openmpi done. If not you
can find the correct replacement for the path
\texttt{/usr/lib64/mpi/gcc/openmpi/bin} 
using \texttt{whereis mpicxx}.
{\small
\begin{lstlisting}
export DEAL_II_DIR =your_mpi_deal_II_version 
export PATH=/usr/lib64/mpi/gcc/openmpi/bin:${PATH}  
export CC=mpicc; export CXX=mpicxx; export FC=mpif90; export FF=mpif77
\end{lstlisting}}
Then compile first the DOpElib library and then the Examples folder with these 
settings. Possibly pay attention that also the deal.II version has been
compiled 
with this MPI path.

\begin{remark}
Be careful that you work in the same shell when only using `export'. 
Or include the export lines into the \texttt{\${HOME}/.bash\_rc} 
\texttt{\${HOME}/.profile}
file to assert that the correct
compiler is always used.
\end{remark}
